---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# ingestr

[![Travis-CI Build Status](https://travis-ci.org/jpshanno/ingestr.svg?branch=master)](https://travis-ci.org/jpshanno/ingestr)
[![Coverage Status](https://img.shields.io/codecov/c/github/jpshanno/ingestr/master.svg)](https://codecov.io/github/jpshanno/ingestr?branch=master)

_An R package for reading environmental data from raw formats into dataframes._ 

This is project was initiated at the inagural [IMCR Hackathon](https://github.com/IMCR-Hackathon/HackathonCentral).  The end product of this effort will be an R package on CRAN.  The package will primarily deal with reading data from files, though there will be some utilities for initial cleanup of files such as removing blank rows and columns at the end of a CSV file.

The guiding principles of the project are that
1. All sources of environmental-related data should be easy to read directly
2. Reading in data should provide a standard output
3. Header information contained within sensor data files should be stored in a standard, easily readable format
4. Associating imported data with its original source is the first step towards good data provenance records and reproducibility

## Installation

You can install ingestr from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jpshanno/ingestr")
```

## Ingesting Data

The main functionality 

```{r example}
## basic example code
```

### Utilize Information

### Batch Ingests

## Preliminary Clean-up Utilities

## QAQCR
